---
title: "Exploring Earthquakes"
author: "Christopher Prener, Ph.D."
date: "January 13, 2017"
output: html_notebook
---

## Introduction
This notebook details the exploration of a data set of earthquakes in Missouri. We created the shapefile referenced here in `MOEarthquakes.Rmd`.

## Dependencies
The following packages are required for this notebook:

```{r}
library(dplyr)        # data wrangling
library(ggplot2)      # plotting

library(skimr)        # descriptive statistics

library(sf)           # spatial data tool
library(tigris)       # download census shapefiles

library(here)         # project organization
library(knitr)        # project documentation
```

## Project Set Up
Before proceeding, we'll make sure our notebook is set up to work with our project data structure.

```{r setup}
opts_knit$set(root.dir = here())
```

See `MOEarthquakes.Rmd` for an explanation of how this function works.

## Load Earthquake Data

## Initial Exploration of the Data
Magnitude of earthquakes is measured using the [Richter scale](https://en.wikipedia.org/wiki/Richter_magnitude_scale). Earthquakes between 2 and 3.9 are considered minor with damage rarely reported. Earthquakes between 4 and 4.9 are considered light, with some minor damage (like objects falling inside buildings). Earthquakes with magnitudes greater than 5 cause increasingly greater amounts of damage. A lesser understood point is that the Richter scale is logrithmic and not linear, meaning that the power of an earthquake increases dramatically with each whole number increase on the scale.

We'll start by using `skimr` to create a quick summary of the magnitude variable:

```{r}
skim(quakes, mag)
```

The most powerful earthquake in the data was a 4.8 magnitude earthquake, and we can see that the bulk of the observations are clustered on the lefthand side of the distribution (half of all observations are between 2 and 2.5).

We can get a better summary of the distribution of earthquake magnitudes by creating a histogram of their magnitude:

```{r}
ggplot(data = quakes, mapping = aes(mag)) + 
  geom_histogram(bins = 28)
```

We can also look at earthquake frequency over time by creating a year variable and then making another histogram. The year variable is made by extracting the first four characters of the `time` variable, which are always the year in these data.

```{r}
quakes <- mutate(quakes, year = as.integer(substr(time, 1, 4)))
  
ggplot(quakes, mapping = aes(year)) +
  geom_histogram(bins = 45)
```

There does seem to be some patterning, with a peak in the mid-1980s and a recent increase after 2010. 

```{r}
ggplot(quakes, mapping = aes(x = year, y = mag)) +
  geom_point() +
  geom_smooth()
```

There is no real increase or decrease in the typical magnitude of earthquakes over each year in the data set. 